%% Report template, by Christian Wolfgang. 20070322

% Document class
\documentclass[10pt,a4paper,onesided]{article}

% Styrelizeren
\usepackage{styrelizer}

\begin{document}

\renewcommand{\thepage}{\roman{page}}
\renewcommand{\thepage}{\arabic{page}}
\setcounter{page}{1}

\section{Introduction}
Give a formal introduction to the problem/assignment at hand. Short
overview of the work done.
\subsection{Motivation}


\section{The Cell Broadband Engine}
What is the CELLBE!
\subsection{Architecture}
Describe the CELLBE architecture. The specs, the pros the cons.
Cons: Lots of data and few operations...
\subsection{Programming techniques}
Double buffering, simdization and so on.

\section{Parallel Programming Techniques}
Maybe this section isn't necessary???

\section{Python}
\subsection{Why Python}
\subsection{Python Objects}
\subsection{The NumPy Structure}
\subsection{Proposed CBE Enhanced Structure}

\section{Blas / Lapack}

\section{Analysis/Datastructure(Working title)}
What structures are available, which of them are interesting, which of
them can be used? Which structure did we choose?
\subsection{Optimizations/improvements}
Can the structure be optimized in any way, how?

\section{Analysis/Dispatcher(Working title)}
Why is a dispatcher important.
\subsection{The SPE}
Something about having multiple programs on an SPE, how to switch
between them.
\subsection{Optimizations/improvements}
Can the dispatcher be optimized in any way, how?

\section{Improving The System}
May be relevant. Keeping blocks on the SPE. Reverse the direction of
blocks used. More...

\section{Implementation(Working title???)}
\subsection{Python/NumPy part}
\subsection{PPE part/Bag of tasks???}
\subsection{SPE part/dispatcher}

\section{Testing, Performance and Results}
\subsection{Testing}
Validation, show/discuss limitiations(if any), conclusion on testing.
\subsection{Performance and Results}
Graphs, discuss results. Did we achieve what we wanted? Discuss any
optimizations done.

\section{Conclusion}
\subsection{Perspective}
\subsection{Future Work}



\end{document}




