\chapter{Guides}

\section{Dispatcher Guide}

This is a brief guide of how the Dispatcher works.

\subsection{Start Up}

The dispatcher must be loaded into the \SPE{}'s \LS{}. This could be
done with:

\examplecode{code/dispatcherload.c}{Loads an SPE ELF into LS}{lst:dispatcherload}

Each \SPE\ must be assigned a unique identification, starting at zero
and counting upwards. Furthermore, each \SPE\ must be given a seed to
initialize its random generator. This could be done by retrieving the
number of seconds by the function \function{time()}.

Given some defines in the file \function{common\_spu.h}, the sizes of the
memory layout in the \LS\ can be changed. Their default values gives
16KB memory per data block, 88KB for the shader and 12KB per actual
shader.

\examplecode{code/ls_layout.c}{The default values for the LS layout}{lst:lslayout}

\subsection{Operations}

Operations for the dispatcher, is given by the operation structure
\struct{Operation\_t}, see code listing \ref{lst:operation_t}.

\examplecode{code/operation_t.c}{The Operation\_t structure}{lst:operation_t}

\subsection{States}

The dispatcher, when successfully initialized, becomes a state
machine. It implements several types of states, whereas they can be
divided into two subgroups:

\begin{itemize}
\item{Functionality with an actual return values}
\item{Functionality with no return value}
\end{itemize}

Functionality with no returns values, do indeed return something, just
not something that involves the dispatcher, but only the given
shader. An actual return value could be something the \PPE\ should
complete. For example \function{sum}, each \SPE\ sums a specific part
of the array and returns their intermediate results to the \PPE{},
whom combines the results into one.

The main difference is, that the \PPE\ program informs each of the
dispatchers where their private intermediate result should be put in
main memory. Otherwise, they look alike.

In table \ref{tbl:states}, the various states can be seen. 

\begin{table}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
State         & Operation ID  & Upload shader & Run shader & Specific return value & Signals \\
\hline
$0 \to 99$    & $0 \to 99$    & Yes           & Yes       & No                     & Yes \\
$100 \to 199$ & $0 \to 99$    & Yes           & Yes       & Yes                    & Yes \\
$200 \to 299$ & $0 \to 99$    & Yes           & No        & No                     & Yes \\
$300 \to 399$ & $0 \to 99$    & No            & Yes       & No                     & No  \\
$400 \to 499$ & $0 \to 99$    & Yes           & No        & Yes                    & Yes \\
$500 \to 599$ & $0 \to 99$    & No            & Yes       & Yes                    & No  \\
\hline
$1000$        & $\varnothing$ & No            & No        & No                     & Yes \\
\hline
\end{tabular}
\caption{Types of states for dispatcher\label{tbl:states}}
\end{table}

\section{Shader Guide}

\section{NumPyCBE}

\subsection{Installation}

NumPyCBE is installed as easily as NumPy, though currently the
installation must be done in \textit{/home/user/numpycbe}. And the
actual install command is thus, in that specific directory,
\texttt{``python setup.py install --prefix=.''}.  If the special
NumPyCBE 64 block optimized version is wanted, the command
\texttt{``python setup64.py install --prefix=.''}, should be used
instead.

To ensure that NumPyCBE can be loaded into any python script
everywhere, the following addition to the \file{.bashrc} file
required:

\examplecode{code/bashrc}{.bashrc file}{lst:bashrc}


\subsection{Functions}

The functionality implemented in NumPyCBE is identical to
NumPy. However, some of the syntax is a little different. This applies
to general operator overloading. For example the index operator and
plus operator.

Table \ref{tbl:overload}, shows the corresponding NumPyCBE syntax.

\begin{table}
\begin{tabular}{|c|c|c|c|c|}
\hline
Side  & Function   & NumPy Syntax & NumPyCBE syntax   & Notes \\
\hline
%Left & Assignment & var[x]       & setIndex(a,x,r,c) & a is the array, x is the value and r\newline and c are the row and column in the array \\
Left  & Assignment & var[i] = x   & setIndex(a,x,r,c) & \mpage{5}{a is the array, x is the value and r and c are the row and column in the array} \\
\hline
Left  & Assignment & var[i] = x   & SetIndex(a,x,i,j) & \mpage{5}{a is the array, x is the value and i and j are the column and row in the array} \\
\hline
Right & Indexing   & x = var[i]   & index(a,x,r,c)    & \mpage{5}{a is the array, x is the value and r and c are the row and column in the array} \\
\hline
Right & Indexing   & x = var[i]   & Index(a,x,i,j)    & \mpage{5}{a is the array, x is the value and i and j are the column and row in the array} \\
\hline
Right & Infix      & a op b       & op(a,b)          & \mpage{5}{a and b are ndarrays or one of them is a scalar, and op is for example +, /, or +=} \\
\hline
\end{tabular}
\caption{Operator overloading conversion\label{tbl:overload}}
\end{table}

In table \ref{tbl:numpycbefuncs} specific NumPyCBE functionalities are listed.


\begin{table}
\begin{tabular}{|c|c|}
\hline
Function        & Notes \\
\hline
SetBlockSize(x) & Sets the currently used block size for FHB\\
\hline
Print(a)        & Prints the ndarray\\
\hline
PrintS(a)       & Prints the ndarray with scientific notation\\
\hline
Create(x,y,v)   & Creates an ndarray with dimensions $y \times x$, with default value v\\
\hline
\end{tabular}
\caption{NumPyCBE specific functionality\label{tbl:numpycbefuncs}}
\end{table}

The following lists the implemented NumPy functionality,

\begin{itemize}
\item{add}
\item{arange}
\item{array}
\item{divide}
\item{len}
\item{lessequal}
\item{multiply}
\item{random}
\item{shape}
\item{subtract}
\item{sum}
\item{solve}
\item{zeros}
\end{itemize}

Though it should be noted, that not all arguments are valid. For
example, only multiply supports slicing.

%\chapter{Test Programs}
%\prepchapter
%\label{chapter:code}

%\section{Monte Carlo Python Code}
%\label{sec:mc_code}

%\subsection{NumPy}
%\examplecode{../code/numpycbe/testprograms/MonteCarlo_org.py}{setup.py file for NumPyCBE}{lst:setup}

%\subsection{NumPyCBE}

%\section{SOR Python Code}
%\label{sec:sor_code}

%\subsection{NumPy}
%\subsection{NumPyCBE}

%\section{Solve Python Code}
%\label{sec:solve_code}

%\subsection{NumPy}
%\subsection{NumPyCBE}









\chapter{Code Listings}
\prepchapter
\label{app:listings}

\section{Shader Code}

\subsection{Generic Binary Function Framework}
\label{app:gbff}
\examplecode{code/genericbinaryshader.c}{Framework for the generic binary shader functionality}{lst:genericshader}


\subsection{Generic Binary Addition}
\label{app:g-addition}
\examplecode{code/genericadd.c}{The \function{\_compute} function for array addition}{lst:genericadd}









\chapter{Output}
\prepchapter
\section{Cell Simulator}

\subsection{STRSM}
\label{app:strsm}

\input{appendix/sim}

\section{SPU Timing Tool}

\subsection{STRSM}
\label{app:strsmtime}

\input{appendix/sputiming}
